Variational Inference for Inverse Reinforcement Learning with Gaussian Processes

### 2018-12-05 mprop/mprop.tex
Proposal is ready for submission, but the submission system is not ready for the proposal

\documentclass{mprop}
\usepackage{graphicx}
\usepackage{amssymb,amsmath,amsthm,mathtools}
\usepackage{url}
\usepackage{bm}
\usepackage{bbm}
\usepackage{booktabs}
\usepackage[multiple]{footmisc}
%\usepackage{fancyvrb}
%\usepackage[final]{pdfpages}

@@

\begin{document}
\title{Variational Inference for Inverse Reinforcement Learning with Gaussian
  Processes}
\author{Paulius Dilkas}
\date{Date of submission placed here} % TODO
\maketitle
\tableofcontents
\newpage

@@

%Imagine a scenario where the agent is driving on a three-lane highway. The
%features of the environment include the agent's speed, lane number, and
%various distances to different types of vehicles in front of and behind the
%agent. The agent can regulate its speed and position. The main objective is, of
%course, to not crash into other vehicles, but one could also add objectives such
%as staying within the speed limit when close to a police vehicle.

Inverse reinforcement learning (IRL)---a problem proposed by Russell in
1998 \cite{DBLP:conf/colt/Russell98}---asks us to find a reward function for a
Markov decision process that best explains a set of given demonstrations. IRL is
important because reward functions can be hard to define manually
\cite{DBLP:conf/icml/PieterN04,DBLP:journals/corr/abs-1806-06877}, and rewards
are not entirely specific to a given environment, allowing one to reuse the same
reward structure in previously unseen environments
\cite{DBLP:journals/corr/abs-1806-06877,DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.
Moreover, IRL has seen a wide array of applications in autonomous vehicle
control \cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} and
learning to predict another agent's behaviour
\cite{DBLP:journals/ai/BogertD18,DBLP:conf/aaai/VogelRGR12,ziebart2008maximum,DBLP:conf/huc/ZiebartMDB08,DBLP:conf/iros/ZiebartRGMPBHDS09}.
Most approaches in the literature (see Section \ref{literature}) make a
convenient yet unjustified assumption that the reward function can be expressed
as a linear combination of features. One proven way to abandon this assumption
is by representing the reward function as a Gaussian process (GP)
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11,DBLP:journals/corr/abs-1208-2112}.
Our main goal is employ variational inference (VI) for recovering the reward
function, which can prove useful in two ways:
\begin{enumerate}
\item As VI tends to be faster than Markov chain Monte Carlo
  sampling \cite{blei2017variational}, we should be able to handle more data.
\item Modelling full posterior distributions for various parameters can result
  in more precise reward predictions, as the model simply holds more
  information.
\end{enumerate}

@@

\begin{definition}
  A \emph{Markov decision process} (MDP) is a set $\mathcal{M} = \{ \mathcal{S},
  \mathcal{A}, \mathcal{T}, \gamma, r \}$, where $\mathcal{S}$ and
  $\mathcal{A}$ are sets of states and actions, respectively; $\mathcal{T} :
  \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is a function
  defined so that $\mathcal{T}(s, a, s')$ is the probability of moving to state $s'$
  after taking action $a$ in state $s$; $\gamma \in [0, 1)$ is the discount
  factor (with higher $\gamma$ values, it makes little difference whether a
  reward is received now or later, while with lower $\gamma$ values the future
  becomes gradually less and less important); and $r : \mathcal{S} \to
  \mathbb{R}$ is the reward function.
\end{definition}

@@

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\bm\lambda$, inducing rewards $\mathbf{u}$, and the rewards for all relevant
states $\mathbf{r}$. While the previous paper to consider this IRL model
computed maximum likelihood estimates for $\bm\lambda$ and $\mathbf{u}$, and
made an assumption that $\mathbf{r}$ in \eqref{eq:r} has zero variance
\cite{DBLP:conf/nips/LevinePK11}, we aim to avoid this assumption and use
VI to approximate the full posterior distribution $\posterior$.
\emph{Variational inference} is an approximation technique for probability
densities \cite{blei2017variational}. Let $\approximation$ be our approximating
family of probability distributions for $\posterior$ with its own hyperparameter
vector $\bm\nu$. Then the job of VI algorithms is to optimise $\bm\nu$ in order
to minimise the \emph{Kullback-Leibler} (KL) divergence between the original
probability distribution and our approximation.  KL divergence (asymmetrically)
measures how different the two distributions are, and in this case can be
defined as \cite{blei2017variational}
\[ \begin{split}
    \DKL(\approximation || \posterior) &= \Eq [\log\approximation -
    \log\posterior ] \\
    &= \Eq[\log\approximation - \log\pfull] \\
    &+ \Eq[\log p(\mathcal{D}, \mathbf{X_u})].
  \end{split}
\]
The last term is both hard to compute and constant w.r.t. $\approximation$
\cite{blei2017variational}, so we can remove it from our optimisation objective.
The negation of what remains is often called the \emph{evidence lower bound}
(ELBO) and is defined as\footnote{Throughout the proposal, all integrals should
  be interpreted as definite integrals over the entire sample space.}
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\bm\lambda\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\end{equation}

By considering full probability distributions instead of point estimates---as
long as the approximations are able to capture important features of the
posterior---our predictions are likely to be more accurate and rely on fewer
assumptions. Moreover, we hope to make use of various recent advancements in VI
for both time complexity and approximation distribution fit (see Section
\ref{literature}), making the resulting algorithm competitive both in terms of
running time and model fit. Therefore, we raise the following research
questions\footnote{For each question, we are primarily comparing the original GP
IRL approach \cite{DBLP:conf/nips/LevinePK11} with our variational version.}:
\begin{enumerate}
\item Does VI allow us to make more accurate reward predictions when given the
  same amount of training time?
\item How does the reward prediction accuracy differ between VI and sampling
  approaches when given the same amount of training data?
\item How well does our VI model approximate the real posterior distribution?
\item Is VI better at learning behaviour driven by multiple goals?
\item Is VI more robust at handling mistakes in demonstrations?
\end{enumerate}

@@

Finally, instead of using GPs to model nonlinear reward functions, one can use a
neural network (NN), as demonstrated by Wulfmeier et al.
\cite{wulfmeier2015maximum}. Their approach benefits from constant time
inference and the ability to learn complex features either from already-given
features or even from raw data. The only disadvantage (as demonstrated in the
paper) is that NNs take longer to learn compared to GPs.

@@

In order to properly investigate the difference between variational inference
and maximum likelihood estimation, we keep other parts of the model the same.
Namely, we set the covariance function to a version of the automatic relevance
detection kernel \cite{DBLP:conf/nips/LevinePK11,neal2012bayesian}
\[ k_{\bm\lambda}(\mathbf{x}_i, \mathbf{x}_j) = \lambda_0\exp\left(
    -\frac{1}{2}(\mathbf{x}_i - \mathbf{x}_j)^\intercal\bm\Lambda(\mathbf{x}_i -
    \mathbf{x}_j) - \mathbbm{1}[i \ne j]\sigma^2\Tr(\bm\Lambda) \right), \]
where $\lambda_0$ is the overall `scale' factor for how similar or distant the
states are, $\bm\Lambda = \diag(\lambda_1, \dots, \lambda_d)$ is a diagonal
matrix that determines the relevance of each feature (where $d$ denotes the
number of features), $\mathbbm{1}$ is defined as
\[ \mathbbm{1}[b] = \begin{cases}
    1 & \mbox{if $b$ is true} \\
    0 & \mbox{otherwise},
  \end{cases} \]
and $\sigma^2$ is set to $10^{-2}/2$ (as the original paper noted that
the value makes little difference to the performance of the algorithm
\cite{DBLP:conf/nips/LevinePK11}). Our vector of hyperparameters for the
covariance function is then $\bm\lambda = (\lambda_0, \dots,
\lambda_d)^\intercal$. Similarly, we keep the expression for the prior of
$\bm\lambda$:
\begin{equation} \label{eq:theta_prior}
  p(\bm\lambda | \mathbf{X_u}) = \exp\left( -\frac{1}{2}\Tr(\Kuu^{-2}) -
    \sum_{i=1}^d \log(\lambda_i + 1) \right).
\end{equation}
Defining the prior for $\bm\lambda$ means that we commit ourselves to
a Bayesian treatment of the variable as opposed to, e.g., treating it as a
variational parameter. While the remainder of the proposal will stick with this
assumption, the alternative is certainly worth exploring in its own right, and
will be investigated after the main formulation is implemented.

@@

Lastly,
$\mathbf{t}^\intercal\mathbb{E}[\Kru^\intercal\Kuu^{-1}]\mathbf{m}$ and both
instances of $\mathbb{E}[V_r(s)]$ come from $p(\mathcal{D} | r)$. The first
expected value can be resolved by making $\bm\lambda$ a variational
parameter. The MDP value function requires more effort. We can start by making
the deterministic training conditional assumption (as is commonly done in
previous work \cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}), which
sets the posterior covariance matrix of $\mathbf{r}$ to the zero matrix. Then
the posterior $\mathbf{r} = \Kru^\intercal\Kuu^{-1}\mathbf{u}$. Hence we still
have a (complicated) function of $\bm\lambda$ and $\mathbf{u}$ in the integral
w.r.t. both variables. The integral (or expected value) w.r.t. $\bm\lambda$ can
be eliminated the usual way, i.e., by making $\bm\lambda$ variational. As for
$\mathbf{u}$, we can borrow a trick from the deep GP IRL paper
\cite{DBLP:conf/uai/JinDAS17} and set $q(\mathbf{u}) = \delta(\mathbf{u} -
\mathbf{m})$, where $\delta$ is the Dirac delta function. The expected value
then becomes $\mathbb{E}[V_r(s)] = V_{\Kru^\intercal\Kuu^{-1}\mathbf{m}}(s)$.

We can then take the BBVI algorithm as
the basis for ours, keeping in mind the numerous improvement possibilities
provided both by the BBVI paper \cite{DBLP:conf/aistats/RanganathGB14} and many
others (see Section \ref{literature}).

@@

The work plan can be summarised as follows:
\begin{enumerate}
\item Derive expressions for the derivatives of the ELBO w.r.t. the variational
  parameters. \emph{Deliverables}: full derivations that can be included as
  supplementary material of the final paper, including formal proofs for
  applications of the dominated convergence theorem. \emph{Date}: 15~December.
\item Implement the algorithm as a more specialised version of Algorithm~2 in
  the BBVI paper \cite{DBLP:conf/aistats/RanganathGB14}. As most of the relevant
  IRL algorithms have been implemented in
  MATLAB\footnote{\url{http://graphics.stanford.edu/projects/gpirl/irl_toolkit.zip}}\footnote{\url{https://github.com/jinming99/DGP-IRL}},
  it makes sense to do the same in order to have accurate time-sensitive
  comparisons between the algorithms. \emph{Deliverables}: a working
  implementation that can be integrated into the IRL toolkit by Levine et al.
  \cite{DBLP:conf/nips/LevinePK11}, and the same algorithm written in pseudo
  code for the paper. \emph{Date}: 1~February.
\item Evaluate the algorithm by comparing it to multiple alternatives, as
  detailed in Section \ref{sec:evaluation}. \emph{Deliverables}: multiple plots
  in an SVG or PDF format, along with code that produces them. \emph{Date}:
  1~April.
\item Write/finalise the paper. \emph{Deliverable}: a 14-page paper.
  \emph{Date}: around 23~April, the deadline has not been updated yet.
\end{enumerate}

>>>>

\documentclass{mprop}
\usepackage{graphicx}
\usepackage{amssymb,amsmath,amsthm,mathtools}
\usepackage{url}
\usepackage{bm}
\usepackage{bbm}
\usepackage{booktabs}
\usepackage[multiple]{footmisc}
\usepackage[UKenglish]{babel}
\usepackage[UKenglish]{isodate}
%\usepackage{fancyvrb}
%\usepackage[final]{pdfpages}

@@

\begin{document}
\title{Variational Inference for Inverse Reinforcement Learning with Gaussian
  Processes}
\author{Paulius Dilkas}
\date{4th December 2018}
\maketitle
\tableofcontents
\newpage

@@

Inverse reinforcement learning (IRL)---a problem proposed by Russell in
1998 \cite{DBLP:conf/colt/Russell98}---asks us to find a reward function for a
Markov decision process (MDP) that best explains a set of given demonstrations.
IRL is important because reward functions can be hard to define manually
\cite{DBLP:conf/icml/PieterN04,DBLP:journals/corr/abs-1806-06877}, and rewards
are not entirely specific to a given environment, allowing one to reuse the same
reward structure in previously unseen environments
\cite{DBLP:journals/corr/abs-1806-06877,DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}.
Moreover, IRL has seen a wide array of applications in autonomous vehicle
control \cite{DBLP:journals/ijsr/KimP16,DBLP:journals/ijrr/KretzschmarSSB16} and
learning to predict another agent's behaviour
\cite{DBLP:journals/ai/BogertD18,DBLP:conf/aaai/VogelRGR12,ziebart2008maximum,DBLP:conf/huc/ZiebartMDB08,DBLP:conf/iros/ZiebartRGMPBHDS09}.
Most approaches in the literature (see Section \ref{literature}) make a
convenient yet unjustified assumption that the reward function can be expressed
as a linear combination of features. One proven way to abandon this assumption
is by representing the reward function as a Gaussian process (GP)
\cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11,DBLP:journals/corr/abs-1208-2112}.
The original approach used maximum likelihood estimation
\cite{DBLP:conf/nips/LevinePK11}, while the goal of this project is to use
variational inference (VI) instead, which learns approximate posterior
probability distributions instead of point estimates. This approach can prove
useful in three major ways:
\begin{enumerate}
\item Modelling full posterior distributions for various parameters can result
  in more precise reward predictions, as the model simply holds more
  information.
\item Having variance estimates for rewards can direct our choice in what data
  should be collected next.
\item An approximate Bayesian treatment of many parameters in the model guards
  against overfitting \cite{DBLP:conf/uai/JinDAS17}.
\end{enumerate}

@@

\begin{definition}
  A \emph{Markov decision process} is a set $\mathcal{M} = \{ \mathcal{S},
  \mathcal{A}, \mathcal{T}, \gamma, r \}$, where $\mathcal{S}$ and
  $\mathcal{A}$ are sets of states and actions, respectively; $\mathcal{T} :
  \mathcal{S} \times \mathcal{A} \times \mathcal{S} \to [0, 1]$ is a function
  defined so that $\mathcal{T}(s, a, s')$ is the probability of moving to state $s'$
  after taking action $a$ in state $s$; $\gamma \in [0, 1)$ is the discount
  factor (with higher $\gamma$ values, it makes little difference whether a
  reward is received now or later, while with lower $\gamma$ values the future
  becomes gradually less and less important); and $r : \mathcal{S} \to
  \mathbb{R}$ is the reward function.
\end{definition}

@@

Given this model, data $\mathcal{D}$, and inducing feature matrix
$\mathbf{X_u}$, our goal is then to find optimal values of hyperparameters
$\bm\lambda$, inducing rewards $\mathbf{u}$, and the rewards for all relevant
states $\mathbf{r}$. While the previous paper to consider this IRL model
computed maximum likelihood (ML) estimates for $\bm\lambda$ and $\mathbf{u}$,
and made an assumption that $\mathbf{r}$ in \eqref{eq:r} has zero variance
\cite{DBLP:conf/nips/LevinePK11}, we aim to avoid this assumption and use
VI to approximate the full posterior distribution $\posterior$.
\emph{Variational inference} is an approximation technique for probability
densities \cite{blei2017variational}. Let $\approximation$ be our approximating
family of probability distributions for $\posterior$ with its own hyperparameter
vector $\bm\nu$. Then the job of VI algorithms is to optimise $\bm\nu$ in order
to minimise the \emph{Kullback-Leibler} (KL) divergence between the original
probability distribution and our approximation.  KL divergence (asymmetrically)
measures how different the two distributions are, and in this case can be
defined as \cite{blei2017variational}
\[ \begin{split}
    \DKL(\approximation || \posterior) &= \Eq [\log\approximation -
    \log\posterior ] \\
    &= \Eq[\log\approximation - \log\pfull] \\
    &+ \Eq[\log p(\mathcal{D}, \mathbf{X_u})].
  \end{split}
\]
The last term is both hard to compute and constant w.r.t. $\approximation$
\cite{blei2017variational}, so we can remove it from our optimisation objective.
The negation of what remains is often called the \emph{evidence lower bound}
(ELBO) and is defined as\footnote{Throughout the proposal, all integrals should
  be interpreted as definite integrals over the entire sample space.}
\cite{DBLP:books/lib/Bishop07,blei2017variational}
\begin{equation} \label{eq:elbo}
  \begin{split}
    \mathcal{L} &= \Eq \left[ \log \frac{\pfull}{\approximation}
    \right] \\
    &= \iiint \approximation \log
    \frac{\pfull}{\approximation}\,d\bm\lambda\,d\mathbf{u}\,d\mathbf{r}.
  \end{split}
\end{equation}

By considering full probability distributions instead of point estimates---as
long as the approximations are able to capture important features of the
posterior---our predictions are likely to be more accurate and rely on fewer
assumptions. Moreover, we hope to make use of various recent advancements in VI
for both time complexity and approximation distribution fit (see Section
\ref{literature}), making the resulting algorithm competitive both in terms of
running time and model fit.

The project is primarily concerned with investigating how a VI formulation of
the GP IRL model compares against the original ML approach. Most importantly, we
aim to compare how the two algorithms converge both over time and as the number
of demonstrations increases. It would also be interesting to see how close the
approximation of the posterior distribution is to the real thing---we discuss
ways to do this in Section \ref{sec:evaluation}. Finally, it is reasonable to
conjecture that VI can outperform point estimates when dealing with more
uncertainty, e.g., when experts make mistakes. We can easily investigate this by
adapting how the evaluation data is generated.

@@

Finally, instead of using GPs to model nonlinear reward functions, one can use a
neural network (NN), as demonstrated by Wulfmeier et al.
\cite{wulfmeier2015maximum}. Their approach benefits from constant time
inference and the ability to learn complex features either from already-given
features or even from raw data. The only disadvantage (as demonstrated in the
paper) is that NNs converge slower than GPs (as measured by the number of
demonstrations).

@@

In order to properly investigate the difference between variational inference
and ML estimation, we keep other parts of the model the same.
Namely, we set the covariance function to a version of the automatic relevance
detection kernel \cite{DBLP:conf/nips/LevinePK11,neal2012bayesian}
\[ k_{\bm\lambda}(\mathbf{x}_i, \mathbf{x}_j) = \lambda_0\exp\left(
    -\frac{1}{2}(\mathbf{x}_i - \mathbf{x}_j)^\intercal\bm\Lambda(\mathbf{x}_i -
    \mathbf{x}_j) - \mathbbm{1}[i \ne j]\sigma^2\Tr(\bm\Lambda) \right), \]
where $\lambda_0$ is the overall `scale' factor for how similar or distant the
states are, $\bm\Lambda = \diag(\lambda_1, \dots, \lambda_d)$ is a diagonal
matrix that determines the relevance of each feature (where $d$ denotes the
number of features), $\mathbbm{1}$ is defined as
\[ \mathbbm{1}[b] = \begin{cases}
    1 & \mbox{if $b$ is true} \\
    0 & \mbox{otherwise},
  \end{cases} \]
and $\sigma^2$ is set to $10^{-2}/2$ (as the original paper noted that
the value makes little difference to the performance of the algorithm
\cite{DBLP:conf/nips/LevinePK11}). Our vector of hyperparameters for the
covariance function is then $\bm\lambda = (\lambda_0, \dots,
\lambda_d)^\intercal$. Similarly, we keep the expression for the prior of
$\bm\lambda$:
\begin{equation} \label{eq:theta_prior}
  p(\bm\lambda | \mathbf{X_u}) = \exp\left( -\frac{1}{2}\Tr(\Kuu^{-2}) -
    \sum_{i=1}^d \log(\lambda_i + 1) \right).
\end{equation}
Defining the prior for $\bm\lambda$ means that we commit ourselves to
a Bayesian treatment of the variable as opposed to, e.g., treating it as a
variational parameter. While the remainder of the proposal will stick with this
assumption, the alternative is certainly worth exploring in its own right, and
will be investigated after the main formulation is implemented.

@@

Lastly,
$\mathbf{t}^\intercal\mathbb{E}[\Kru^\intercal\Kuu^{-1}]\mathbf{m}$ and both
instances of $\mathbb{E}[V_r(s)]$ come from $p(\mathcal{D} | r)$. The first
expected value can be resolved by making $\bm\lambda$ a variational
parameter. The MDP value function requires more effort. We can start by making
the deterministic training conditional assumption (as is commonly done in
previous work \cite{DBLP:conf/uai/JinDAS17,DBLP:conf/nips/LevinePK11}), which
sets the posterior covariance matrix of $\mathbf{r}$ to the zero matrix. Then
the posterior $\mathbf{r}$ becomes $\Kru^\intercal\Kuu^{-1}\mathbf{u}$. But we
still have a (complicated) function of $\bm\lambda$ and $\mathbf{u}$ in the
integral w.r.t. both variables. The integral (or expected value) w.r.t.
$\bm\lambda$ can be eliminated the usual way, i.e., by making $\bm\lambda$
variational. As for $\mathbf{u}$, we can borrow a trick from the deep GP IRL
paper \cite{DBLP:conf/uai/JinDAS17} and set $q(\mathbf{u}) = \delta(\mathbf{u} -
\mathbf{m})$, where $\delta$ is the Dirac delta function. This is essentially a
different way to impose a zero-variance restriction, modelling $\mathbf{u}$ as a
random vector with variance tending to zero. The expected value then becomes
$\mathbb{E}[V_r(s)] = V_{\Kru^\intercal\Kuu^{-1}\mathbf{m}}(s)$.

While the ELBO can now be computed without approximations, we ended up imposing
two zero-variance restrictions, resulting in a less informative model. Moreover,
we had to abandon the Bayesian treatment of $\bm\lambda$, making the model more
vulnerable to overfitting.

Assuming that such workarounds will not be needed (or desired), we can then take
the BBVI algorithm as the basis for our algorithm, while keeping in mind the
numerous improvement possibilities provided both by the BBVI paper
\cite{DBLP:conf/aistats/RanganathGB14} and many others (see Section
\ref{literature}).

@@

The work plan can be summarised as follows:
\begin{enumerate}
\item Derive expressions for the derivatives of the ELBO w.r.t. the variational
  parameters. \emph{Deliverables}: full derivations that can be included as
  supplementary material of the final paper, including formal proofs for
  applications of the dominated convergence theorem. \emph{Date}: 15~December.
\item Implement the algorithm as a more specialised version of Algorithm~2 in
  the BBVI paper \cite{DBLP:conf/aistats/RanganathGB14}. As most of the relevant
  IRL algorithms have been implemented in
  MATLAB\footnote{\url{http://graphics.stanford.edu/projects/gpirl/irl_toolkit.zip}}\footnote{\url{https://github.com/jinming99/DGP-IRL}},
  it makes sense to do the same in order to have accurate time-sensitive
  comparisons between the algorithms. \emph{Deliverables}: a working
  implementation that can be integrated into the IRL toolkit by Levine et al.
  \cite{DBLP:conf/nips/LevinePK11}, and the same algorithm written in pseudo
  code for the paper. \emph{Date}: 1~February.
\item Evaluate the algorithm by comparing it to multiple alternatives, as
  detailed in Section \ref{sec:evaluation}. \emph{Deliverables}: multiple plots
  in an SVG or PDF format, along with code that produces them. \emph{Date}:
  1~April.
\item Write/finalise the paper. \emph{Deliverable}: a 14-page paper.
  \emph{Date}: 12~April.
\end{enumerate}
